The state of knowledge for marine invertebrate biodiversity in the continental US

### 2017-07-07 manuscript.Rmd
edit manuscript

Untangling the relative contribution of these factors to explain the discrepencies is difficult as it would require to know the criteria used by the experts who have assembled the lists. However, by querying iDigBio we found that for the `r format(not_really_in_database$idig_gom_n_spp_not_in_db, big.mark=",")` species that do not appear within the geographic range of the Gulf of Mexico but are on the list, `r format_output(not_really_in_database$idig_gom_p_spp_not_in_db*100)`% are truly absent from iDigBio, while for the `r format(not_really_in_database$idig_gom_n_spp_no_coords, big.mark=",")` species that appear in iDigBio, `r format_output(not_really_in_database$idig_gom_p_spp_no_coords*100)`%  do not have geographic coordinates associated with them, and for the ones that do have geographic coordinates, `r format_output(not_really_in_database$idig_gom_p_spp_within_usa*100)`% ($n=$  `r format(not_really_in_database$idig_gom_n_spp_within_usa, big.mark=",")`) fall within the US waters. These numbers indicate that the lack of geographic coordinates associated with the specimens in the database contribute at least significantly to the discrepency, but that a large fraction of these species have no digital records of their occurrence in the Gulf of Mexico. In other words, additional digitization of collections (including capturing their geographic coordinates information) are needed. For instance, one can expect that the digitization coverage for the Gulf of Mexico will greatly increase once records from the Florida Wildlife Research Institute collections (St Petersburg, FL) become integrated into the biodiversity information systems.

@@

We combined `r format(nrow(idigbio_records), big.mark=",")` museum records for marine invertebrates collected within the EEZ of the continental US representing `r length(unique(tolower(idigbio_records[["data.dwc:phylum"]])))` phyla and `r format(length(unique(idigbio_records$worms_valid_name)), big.mark=",")` species. Most samples were collected on the continental shelf (Fig. \@ref(fig:map-diversity-sampling)A). For the East coast, species richness is the highest in South Florida but it is relatively uniform along the coastline. For the West coast, the Pacific Northwest, the San Francisco Bay Area, and Southern California have the highest richness (Fig. \@ref(fig:map-diversity-sampling)B).

@@

These patterns of spatial distribution of the species richness are strongly confounded by the sampling effort. When investigating the relation between the number of samples and the number of species for each cell of the map (Fig. \@ref(fig:correlation-sampling)), it takes about 100 records per cell before there is some redundancy in the sampled species (each record represent a different species).

```{r correlation-sampling, fig.cap='Relation between number of records and number of species for each cell of the map used in Fig. \\@ref(fig:map-diversity-sampling). (\\#fig:correlation-sampling)'}
knitr::include_graphics("figures/plot_sampling_effort.pdf")
```

The inadequacy of the sampling effort compared to the diversity of the fauna is also reflected by the fact that `r prop_singleton_species`% of the species are known from a single record. The strong correlation between the number of samples collected and the species richness suggests that any new sampling effort will lead to new observations and new species (Fig. \@ref(fig:correlation-sampling)). Another indication of this inadequacy is the relation between the number of collected specimen and the number of species recorded through time (Fig. \@ref(fig:idigbio-through-time)). Initial explorations in the late ninetieth century lead to an initial exponential growth in the knowledge of the fauna, that tapered off, until the 1970's when intensive collecting efforts led to an increase in the recording of new taxa (Fig. \@ref(fig:idigbio-through-time)). Since then, the slowing down of collections after the 1970s has been accompanied with a slowing down in the discovery of new records: `r n_spp_comparison$n_spp_2017_2000` species have been documented for the first time since 2000 (`r format_output(n_spp_comparison$n_spp_2017_2000/18)` species.year^-1^), while `r n_spp_comparison$n_spp_1970_1960` have been documented between 1960 and 1970 (`r format_output(n_spp_comparison$n_spp_1970_1960/11)` species.year^-1^). Therefore, coordinated large-scale biodiversity expeditions will continue to generate a wealth of data and new records. For instance, a recent study in Western Australia reported that 135 samples led to 805 species of crustaceans and polychaetes, 94.6% and 72% of which were undescribed [@Poore2014].

```{r idigbio-through-time, fig.cap='Cumulative number of samples (A) and species (B) of marine invertebrates recorded through time based on iDigBio records. Relation between number of samples collected, and number of species recorded for the first time (C). (\\#fig:idigbio-through-time)'}
knitr::include_graphics("figures/idigbio_through_time.pdf")
```

Predicting the geographic range of species, and how they can be impacted by global change has undergone important advances recently. However, it is generally recognized that to obtain reliable results such efforts require to have accurate geographic information for a minimum of 10 specimens but having at least 30 is preferred. Based on iDigBio data, `r prop_10_specimens_species`% of species have less than 10 specimens and `r prop_30_specimens_species`% have less than 30 specimens. This suggests that our understanding of the geographic range of species is limited as most species are known from only a few locations, while efforts to estimate their current and future ranges based on occurrence data are rendered moot by the low number of data points available.

@@

To assess the status of taxonomic expertise in the US, we calculated the proportion of specimens identified at the species level through time. For all taxa, excepts for molluscs, there is a consistent decline, indicating that specimens remain for the most part unidentified in collections (Fig.  \@ref(fig:identification-through-time)). For diverse groups such as annelids or arthropods, the proportion of specimens identified at the species-level is now below 30%. The number of scientists describing new species might be increasing globally [@Costello2013], but local experts are needed to keep identifying the species that get collected.

@@

Since its introduction in the early 2000's, DNA barcoding, the sequencing of a common locus across taxa bearing species-specific information, has revitalized taxonomy by facilitating species delineation [@Meyer2005-plos]. More recently, advances in high-throughput sequencing technologies have enabled the parallel sequencing of thousands of species that occur in environmental samples cost effectively [@Leray2015]. This approach opens up new research venues as it becomes possible to estimate the diversity from species-rich samples that were virtually inaccessible before. This is particularly true in marine communities that include multiple phyla and organisms varying in size by order of magnitude, making identification challenging. However, the utility of the data generated by these projects is enhanced by the possibility of matching DNA sequences obtained to previously sequenced and correctly identified species [@Leray2015]. With decreasing sequencing costs, and the prospect of automating sample collection, these technologies may soon be used routinely to monitor biodiversity at defined temporal frequency [@Bohan2017]. DNA barcoding provides the possibility to link these newly generated data with historical natural history knowledge about species and their distributions. However, the insights we will gain from these approaches are only going to be as good as the level of completeness and accuracy of the information included in reference databases.

@@

The proportion of species that have DNA barcodes available is relatively constant across the better studied phyla at the scale of the EEZ. It varies between 12% for bryozoans to 28% for cnidarians. The proportions for species known to occur in the Gulf of Mexico are usually similar than for the US, while typically a higher fraction of species from the Pacific Northwest are available. In 2012, it was estimated that a global scale, less than 3% of the marine species had been sequenced [@Appeltans2012], and that this number was even lower in some areas (e.g. 2.6% for the Antarctic [@Grant2009]). This suggests that even as many additional species have had their DNA barcodes made available since these studies, the US fauna is generally better characterized than other parts of the world.

@@

```{r proportion-barcoded, fig.cap='Proportion of species known to occur in the US EEZ based on iDigBio records, in the Gulf of Mexico [@Tunnell2009], in the Pacific Northwest[@Kozloff1987], that have publicly available DNA barcodes. Numbers next to the bars indicate the number of species.'}
knitr::include_graphics("figures/proportion_barcoded.pdf")
```

@@

Similarly, a significant fraction of museum records do not have or have inaccurate geographic information associated with them. While tools have been developed to convert names of locales into GPS coordinates on land [@rios2010geolocate], similar approaches may not work for locales at sea. However, geographic information is available in other forms, and can be digitized.

For both the spelling of scientific names, and the geographic information, computing tools can be developed to detect issues and propose solutions. There are opportunities for services using publicly available data to be developed to make curators and collection managers aware of these issues so that they can be fixed at the source. It however requires that resources are made available to museum collections to address these issues.

@@

<!-- - develop infrastructure and tools to continuously check and validate data quality in databases. -->

@@

This study highlights that we have too little data about too few species of marine invertebrates, we recommend that additional coordinated field expeditions are organized to sample more thoroughly the marine environment. The relation between sampling effort and diversity from museum collections makes clear that any additional samples will improve greatly our knowledge of marine biodiversity. Concerted and collaborative efforts are needed to share expertise and limited resources.

@@

Integrating metabarcoding as part of the data collection conducted in biodiversity surveys is a promising approach to screen an habitat for its potential to lead to new species or species that have not yet public DNA barcodes. For instance, by analyzing metabarcoding samples that contain planktonic larvae from a given habitat, we could get an indication of which taxonomic group would benefit most from additional sampling effort or give an indication of the proportion of species that has not been sampled yet. Additionally, data generated by metabarcoding efforts can be re-interpreted as more DNA barcodes associated with correctly identified specimens become available. Building the library of DNA barcodes in parallel of metabarcoding efforts ensures that database matches correspond to actual specimens deposited in natural history collections, ruling out possible artifacts of the method. With such approaches, the geographic distribution of species can build on and linked to existing records rather than being limited to matching DNA sequences across studies.

@@

Even if technological advances can greatly speed up the documentation of species and their distributions, taxonomists and experts are needed  to collect, sort, process, and identify samples. Taxonomic expertise within the US is dwindling, and there are now only a handful of US natural history museums with active marine invertebrates collections, and curatorial staff.

Crowd-sourced data and citizen science initiatives have been proposed to augment data about species abundance and distributions [@Pimm2014]. For marine taxa, beyond the application to near-shore habitats on easily recognizable species [@Delanay2008], large-scale efforts are currently not possible given the specialized equipment needed to sample most marine habitats and the expertise required to identify species. There however opportunities to utilize citizen science to improve biodiversity data and engage the general public into marine conservation issues. For instance it has been successfully used to classify plankton images [@Robinson2017], transcribe hand-written museum labels [@Hill2012], or be involved in bioblitzes [@Ballard2017]. These initiatives however involve a lot less people compared to projects related to birds.

>>>>

Untangling the relative contribution of these factors to explain the discrepencies is difficult as it would require to know the criteria used by the experts who have assembled the lists. However, by querying iDigBio we found that for the `r format(not_really_in_database$idig_gom_n_spp_not_in_db, big.mark=",")` species that do not appear within the geographic range of the Gulf of Mexico but are on the list, `r format_output(not_really_in_database$idig_gom_p_spp_not_in_db*100)`% are truly absent from iDigBio, while for the `r format(not_really_in_database$idig_gom_n_spp_no_coords, big.mark=",")` species that appear in iDigBio, `r format_output(not_really_in_database$idig_gom_p_spp_no_coords*100)`%  do not have geographic coordinates associated with them, and for the ones that do have geographic coordinates, `r format_output(not_really_in_database$idig_gom_p_spp_within_usa*100)`% ($n=$  `r format(not_really_in_database$idig_gom_n_spp_within_usa, big.mark=",")`) fall within the US waters. These numbers indicate that the lack of geographic coordinates associated with the specimens in the database contribute at least significantly to the discrepency, but that a large fraction of these species have no digital records of their occurrence in the Gulf of Mexico. In other words, additional digitization of collections (including capturing their geographic coordinates information) are needed. For instance, one can expect that the digitization coverage for the Gulf of Mexico will greatly increase once records from the Florida Wildlife Research Institute collections (St Petersburg, FL) become integrated into biodiversity information systems.

@@

We combined `r format(nrow(idigbio_records), big.mark=",")` museum records for marine invertebrates collected within the EEZ of the continental US representing `r length(unique(tolower(idigbio_records[["data.dwc:phylum"]])))` phyla and `r format(length(unique(idigbio_records$worms_valid_name)), big.mark=",")` species from iDigBio. Most samples were collected on the continental shelf (Fig. \@ref(fig:map-diversity-sampling)A). For the East coast, species richness is the highest in South Florida but it is relatively uniform along the coastline. For the West coast, the Pacific Northwest, the San Francisco Bay Area, and Southern California have the highest richness (Fig. \@ref(fig:map-diversity-sampling)B).

@@

These patterns of spatial distribution for the species richness are strongly confounded by the sampling effort. Investigating the relation between the number of samples and the number of species for each cell of the map (Fig. \@ref(fig:correlation-sampling)) reveals that it takes about 100 records per cell before there is clear redundancy in the sampled species (each record represent a different species).

```{r correlation-sampling, fig.cap='Relation between number of records and number of species for each cell of the map used in Fig. \\@ref(fig:map-diversity-sampling). The black line represents the 1:1 relation. (\\#fig:correlation-sampling)'}
knitr::include_graphics("figures/plot_sampling_effort.pdf")
```

The inadequacy of the sampling effort compared to the diversity of the fauna is also reflected by the fact that `r prop_singleton_species`% of the species are known from a single record. Another indication of this inadequacy is the relation between the number of collected specimen and the number of species recorded through time (Fig. \@ref(fig:idigbio-through-time)). Initial explorations in the late ninetieth century lead to an initial exponential growth in the knowledge of the fauna, that tapered off, until the 1970's when intensive collecting efforts led to an increase in the recording of new taxa (Fig. \@ref(fig:idigbio-through-time)). Since then, the slowing down of collections after the 1970s has been accompanied with a slowing down in the discovery of new records: `r n_spp_comparison$n_spp_2017_2000` species have been documented for the first time since 2000 (`r format_output(n_spp_comparison$n_spp_2017_2000/18)` species.year^-1^), while `r n_spp_comparison$n_spp_1970_1960` have been documented between 1960 and 1970 (`r format_output(n_spp_comparison$n_spp_1970_1960/11)` species.year^-1^). The strong correlations between (1) the number of samples collected and the species richness (Fig. \@ref(fig:correlation-sampling)).; (2) the number of specimens collected per year and the number of species recorded for the first time (Fig. \@ref(fig:idigbio-through-time)C), suggest that any new sampling effort will lead to new observations and new species. Coordinated large-scale biodiversity expeditions continue to be needed to improve our understanding of the diversity of species and their spatial distributions. While less explored than the US, a recent study in Western Australia reported that 135 samples led to 805 species of crustaceans and polychaetes, 94.6% and 72% of which were undescribed [@Poore2014], highlighting the potential for discovery in the oceans.

```{r idigbio-through-time, fig.cap='Cumulative number of samples (A) and species (B) of marine invertebrates recorded through time based on iDigBio records. Relation between number of samples collected, and number of species recorded for the first time (C), each point represents numbers for a given year. (\\#fig:idigbio-through-time)'}
knitr::include_graphics("figures/idigbio_through_time.pdf")
```

Predicting the geographic range of species, and how they will be impacted by global change has undergone important advances recently [@xxxx]. However, it is generally recognized that to obtain reliable results such efforts require to have accurate geographic information for a minimum of 10 specimens but having at least 30 is preferred [@xxxx]. Based on iDigBio data, `r prop_10_specimens_species`% of species have less than 10 specimens and `r prop_30_specimens_species`% have less than 30 specimens. This suggests that our understanding of the geographic range of species is limited as most species are known from only a few locations, while efforts to estimate their current and future ranges based on occurrence data are rendered moot by the low number of data points available.

@@

To assess the status of taxonomic expertise in the US, we calculated the proportion of specimens identified at the species level through time. For all taxa, excepts for molluscs, there is a consistent decline, indicating that specimens collected recently remain for the most part unidentified in collections (Fig.  \@ref(fig:identification-through-time)). For diverse groups such as annelids or arthropods, the proportion of specimens identified at the species-level is now below 30%. The number of scientists describing new species might be increasing globally [@Costello2013], but taxonomic experts continue to be needed to identify specimens recently collected.

@@

Since its introduction in the early 2000's, DNA barcoding, the sequencing of a common locus across taxa bearing species-specific information, has revitalized taxonomy by facilitating species delineation [@Meyer2005-plos]. More recently, advances in high-throughput sequencing technologies have enabled the parallel sequencing of thousands of species that occur in environmental samples cost effectively [@Leray2015]. This approach opens up new research venues as it becomes possible to estimate the diversity from species-rich samples that were virtually inaccessible before. This is particularly true in marine communities that include multiple phyla and organisms varying in size by orders of magnitude, making identification challenging. However, the utility of the data generated by these projects is enhanced by the possibility of matching DNA sequences obtained to previously sequenced and correctly identified species [@Leray2015]. With decreasing sequencing costs, and the prospect of automating sample collection, these technologies may soon be used routinely to monitor biodiversity continuously [@Bohan2017]. DNA barcoding provides the possibility to link these newly generated data with historical natural history knowledge about species and their distributions. However, the insights we will gain from these approaches are only going to be as good as the level of completeness and accuracy of the information included in reference databases.

@@

The proportion of species that have DNA barcodes available is relatively constant across the better studied phyla at the scale of the EEZ. It varies between 12% for bryozoans to 28% for cnidarians (\@ref(fig:proportion-barcoded)). The proportions for species known to occur in the Gulf of Mexico are usually similar than for the US, while typically a higher fraction of species from the Pacific Northwest are available. In 2012, it was estimated that, at a global scale, less than 3% of the marine species had been sequenced [@Appeltans2012], and that this number was even lower in some areas (e.g. 2.6% for the Antarctic [@Grant2009]). This suggests that even as many additional species have had their DNA barcodes made available since these studies, the US fauna is generally better characterized than other parts of the world.

@@

```{r proportion-barcoded, fig.cap='Proportion of species known to occur in the US EEZ based on iDigBio records, in the Gulf of Mexico [@Tunnell2009], in the Pacific Northwest[@Kozloff1987], that have publicly available DNA barcodes. Numbers next to the bars indicate the number of species for which DNA barcodes are available. (\\#fig:proportion-barcoded)'}
knitr::include_graphics("figures/proportion_barcoded.pdf")
```

@@

Similarly, a significant fraction of museum records have missing or inaccurate geographic information associated with them. While tools have been developed to convert names of locales into GPS coordinates on land [@rios2010geolocate], similar approaches may not work for locales at sea. However, geographic information is available in other forms, and can be digitized.

For both the spelling of scientific names, and the geographic information, computing tools can be developed to detect issues and propose solutions. There are opportunities to develop computing services that use publicly available data to flag and fix these issues directly in collection databases. It however requires that resources are made available to museum collections to address these issues.

@@

@@

This study highlights that we have too little data about too few species of marine invertebrates. We recommend that additional coordinated field expeditions are organized to sample more thoroughly the marine environment. The relation between sampling effort and diversity from museum collections makes clear that any additional samples will improve greatly our knowledge of marine biodiversity. Concerted and collaborative efforts are needed to share expertise and limited resources.

@@

Integrating metabarcoding as part of the data collection conducted in biodiversity surveys is a promising approach to screen an habitat for its potential to lead to new discoveries. For instance, by analyzing metabarcoding samples that contain planktonic larvae from a given habitat, we could get an indication of which taxonomic group would benefit most from additional sampling effort or give an indication of the proportion of species that has not been sampled yet. Additionally, data generated by metabarcoding efforts can be re-interpreted as more DNA barcodes associated with correctly identified specimens become available. Building the library of DNA barcodes in parallel of metabarcoding efforts ensures that database matches correspond to actual specimens deposited in natural history collections, ruling out possible artifacts of the method. With such approaches, the geographic distribution of species can build on and be linked to existing records rather than being limited to matching DNA sequences across studies.

@@

Even if technological advances can greatly speed up the documentation of species and their distributions, taxonomists and experts are needed  to collect, sort, process, and identify samples. Taxonomic expertise within the US is dwindling, and there are now only a handful of US natural history museums with active marine invertebrates collections.

Crowd-sourced data and citizen science initiatives have been proposed to augment data about species abundance and distributions [@Pimm2014]. For marine taxa, beyond the application to near-shore habitats on easily recognizable species [@Delaney2008], large-scale efforts are currently not possible given the specialized equipment needed to sample most marine habitats and the expertise required to identify species. There are however opportunities for citizen science to improve biodiversity data, educate and engage the general public with marine conservation issues. For instance it has been successfully used to classify plankton images [@Robinson2017], transcribe hand-written museum labels [@Hill2012], or be involved in bioblitzes [@Ballard2017]. These initiatives however involve a lot less people compared to projects related to birds.
